Online learning methods for financial time series forecasting

### 2017-11-26 slides/defense-slides.tex
more images

\documentclass{beamer}
\usetheme{Warsaw}
%\usetheme{metropolis}
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{adjustbox}
\usepackage{color}
\usepackage{tikz}
\usepackage{array}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

   \definecolor{myblue1}{RGB}{35,119,189}
    \definecolor{myblue2}{RGB}{95,179,238}
    \definecolor{myblue3}{RGB}{129,168,207}
    \definecolor{myblue4}{RGB}{26,89,142}

    \setbeamercolor*{structure}{fg=myblue1,bg=blue}
    \setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
    \setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
    \setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
    \setbeamercolor*{palette quaternary}{fg=black,bg=white}

    \setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
    \setbeamercolor*{block title example}{fg=white,bg=myblue4}
    \setbeamercolor*{frametitle}{fg=black}

    \setbeamertemplate{blocks}[rounded][shadow=true]

    \makeatletter
    \pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
      color(0pt)=(myblue2);
      color(\paperwidth)=(white)}

    \defbeamertemplate*{footline}{mysplit theme}
    {%
      \leavevmode%
      \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
        \usebeamerfont{author in head/foot}\insertshortauthor
      \end{beamercolorbox}%
      \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
        \usebeamerfont{title in head/foot}\insertshorttitle\hfill
        \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
      \end{beamercolorbox}}%
      \vskip0pt%
    }
    \makeatother

@@

\titlegraphic{
    \tikz[overlay,remember picture]
        \node[at=(current page.south), anchor=south] {
            \includegraphics[height=1.5cm]{img/DIT}
        };
}

\begin{document}
\begin{frame}[plain]
\titlepage
\end{frame}

@@

A weak stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

@@

\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}

@@

\begin{frame}
\frametitle{$\Omega$ matrix properties}
The matrix $\Omega$ has the following properties:
\begin{itemize}
\item If $\Omega = 0$ there is no cointegration
\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
I(1) but stationary
\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
and the matrix $\Omega$ can be expressed as $\Omega =
\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
matrices and $rank(\alpha)=rank(\beta)=r$.
\end{itemize}
The columns of $\beta$ contains the cointegration vectors and the rows of
$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
procedure whereas $\alpha$ has to be determined as a
variable in the VECM.
\end{frame}

\begin{frame}
\frametitle{VECM limitations}

The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data  mainly due to computational limitations:

\begin{itemize}
\item Calculation of the cointegration vectors
\item VECM parameters are obtained using the ordinary least squares (OLS) method.
\end{itemize}

\section{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{Proposal}
To adapt VECM to be used with high frequency data, where the best parameters are updated when new data.
\end{block}

\begin{block}{Hypothesis}
An algorithm based on cointegration and high
performance computing will allow faster forecasting
algorithms for financial time series to be obtained while maintaining good accuracy levels.
\end{block}
\end{frame}

\section{First Contribution}
\begin{frame}
\frametitle{First approach}
To use VECM with high frequency data, where optimised parameters are obtained when new data. The optimisation was made in a distributed environment.
\end{frame}

\subsection{Data}
\begin{frame}
\frametitle{Data}
\begin{itemize}
\item Tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
\item This data was collected from Dukascopy, a free
database which gives access to the Swiss Foreign Exchange marketplace.
\item The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. 
%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Cointegration vectors distribution}
  \begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}
\end{frame}

@@

\subsection{Parameters}
\begin{frame}
\frametitle{Parameters optimization}
VECM order and ARIMA parameters were selected using AIC.

\begin{table}[ht]
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}

>>>>

\documentclass{beamer}
\usetheme{Warsaw}
%\usecolortheme{lily}
%\usetheme{metropolis}
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{tikz}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

%   \definecolor{myblue1}{RGB}{35,119,189}
%    \definecolor{myblue2}{RGB}{95,179,238}
%    \definecolor{myblue3}{RGB}{129,168,207}
%    \definecolor{myblue4}{RGB}{26,89,142}
%
%    \setbeamercolor*{structure}{fg=myblue1,bg=blue}
%    \setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
%    \setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
%    \setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
%    \setbeamercolor*{palette quaternary}{fg=black,bg=white}
%
%    \setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
%    \setbeamercolor*{block title example}{fg=white,bg=myblue4}
%    \setbeamercolor*{frametitle}{fg=black}
%
%    \setbeamertemplate{blocks}[rounded][shadow=true]
%
%    \makeatletter
%    \pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
%      color(0pt)=(myblue2);
%      color(\paperwidth)=(white)}
%
%    \defbeamertemplate*{footline}{mysplit theme}
%    {%
%      \leavevmode%
%      \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
%        \usebeamerfont{author in head/foot}\insertshortauthor
%      \end{beamercolorbox}%
%      \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
%        \usebeamerfont{title in head/foot}\insertshorttitle\hfill
%        \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
%      \end{beamercolorbox}}%
%      \vskip0pt%
%    }
%    \makeatother

\usebackgroundtemplate{%             declare it
\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
}

@@

\titlegraphic{
    \tikz[overlay,remember picture]
        \node[at=(current page.south east), anchor=south east] {
            \includegraphics[height=1.5cm]{img/DIT}
        };
}

\begin{document}

\begin{frame}[plain]
\titlepage
\end{frame}
\usebackgroundtemplate{ }

@@

A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

@@

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
The columns of $\beta$ contains the cointegration vectors and the rows of
$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
procedure whereas $\alpha$ has to be determined as a
variable in the VECM.

@@

\begin{frame}
\frametitle{$\Omega$ matrix properties}
The matrix $\Omega$ has the following properties:
\begin{itemize}
\item If $\Omega = 0$ there is no cointegration
\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
I(1) but stationary
\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
and the matrix $\Omega$ can be expressed as $\Omega =
\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
matrices and $rank(\alpha)=rank(\beta)=r$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{VECM limitations}
The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data  mainly due to computational limitations:
\begin{itemize}
\item Calculation of the cointegration vectors
\item VECM parameters are obtained using the ordinary least squares (OLS) method.
\end{itemize}
\end{frame}

\section{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{Proposal}
To adapt VECM to be used with high frequency data, where the best parameters are updated when new data arrives.
\end{block}
\begin{block}{Hypothesis}
An algorithm based on cointegration and high performance computing will allow faster forecasting
algorithms for financial time series to be obtained while maintaining good accuracy levels.
\end{block}
\end{frame}

\section{First Contribution}
\begin{frame}
\frametitle{First approach}
An adaptive Vector Error Correction algorithm that:
\begin{itemize}
\item considers the last most recent data (windows size L)
\item finds optimal parameters of VECM in that window.
\item the optimisation is made in a distributed environment.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{First approach idea}
\begin{figure}
\includegraphics[width=0.65\textwidth]{img/slidingwindow}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{How to choose L?}
  \begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}
\end{frame}

\subsection{Data}
\begin{frame}
\frametitle{Data}
\begin{itemize}
\item Tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
\item This data was collected from Dukascopy, a free
database which gives access to the Swiss Foreign Exchange marketplace.
\item The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. 
%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{itemize}
 \begin{figure}[!h]
    \includegraphics[width=0.2\textwidth]{img/dukascopy}
 \end{figure}
\end{frame}

@@

\subsection{Parameters}
\begin{frame}
\frametitle{Parameters optimization}
VECM order and ARIMA parameters were selected using AIC.
\begin{table}[ht]
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}
